\textbf{Linear Algebra -- Part II}\\A summary for MIT 18.06SC

### 2020-07-10 skills/math/linear_algebra/linear_algebra_summary_p2.tex
linear_algebra

\subsection{Orthogonal vectors and subspaces}
\begin{defn}[\textit{Orthogonal vectors}]\label{def:vortho}
	Suppose two vectors $x,y \in \bbR^{n}$. The vectors $x$ and $y$ are orthogonal \textbf{iff} $x^T y = y^T x = 0$, denoted $x \perp y$.
\end{defn}

\begin{defn}[\textit{Orthogonal subspaces}]\label{def:sortho}
	Suppose two subspaces $S,T$. The subspaces $S$ and $T$ are orthogonal \textbf{iff} for any $s \in S$ and for any $ t \in T$, $s^T t = t^T s = 0$, denoted $S \perp T$.
\end{defn}

Given a matrix $\mA \in \bbR^{m \times n}$, there are four subspaces related to $\mA$: column space $C(\mA)$, row space $C(\mA^T)$, nullspace $N(\mA)$, and left nullspace $N(\mA^T)$. Suppose the matrix rank of $\mA$ is $rank(\mA) = r$, the dimensions of these subspaces are: 
\[ dim(C(\mA)) = dim(C(\mA^T)) = r,\quad  dim(N(A)) = n-r,\quad dim(N(\mA^T)) = m-r. \]

\begin{thm}[Orthogonality of matrix subspaces]\label{thm:ortho}
	Suppose a matrix $\mA \in \bbR^{m \times n}$. The row space $C(\mA^T)$ and the nullspace $N(\mA)$ are orthogonal; the column space $C(\mA)$ and the left nullspace $N(\mA^T)$ are orthogonal i.e.\
	\[ C(\mA^T) \perp N(\mA); \quad C(\mA) \perp N(\mA^T).  \]
\end{thm}

\begin{proof}
	Consider the matrix $\mA \in \bbR^{m \times n}$. For any vector $x \in N(\mA)$, we have $\mA x = 0$. 
	\begin{align}
		\mA x  = \begin{bmatrix}
			a_1^T x \\ \vdots \\ a_m^T x 
		\end{bmatrix} = \begin{bmatrix}
			0\\ \vdots \\0
		\end{bmatrix},
	\end{align}
	where $a_i \in \bbR^n, i \in [m]$ is the $i$-th  row of $\mA$. By the definition~\ref{def:vortho}, $x$ is orthogonal with the rows in matrix $\mA$. For any vector $v \in C(\mA^T)$, $v$ is a linear combination of the rows, i.e.\ $v = c_1 a_1 + ...+c_m a_m$, where $c_i, i\in [m]$ are constants. Multiplying vectors $v$ and $x$,
	\[  v^T x =  c_1 a_1^T x + \cdots + c_m a_m^T x = 0.  \]
	Therefore, $v \perp x$, and $N(\mA) \perp C(\mA^T)$.

@@

\begin{thm}[Relationship between $\mA^T \mA$ and $\mA$]\label{thm:aa}
	Consider a matrix $\mA \in \bbR^{m \times}$. We have
	\begin{equation}
		N(\mA^T \mA) = N(\mA) ; \quad rank(\mA^T \mA) = rank(\mA).
	\end{equation}
\end{thm}
\begin{proof}
	Consider the matrix $\mA \in \bbR^{m \times n}$.

	If $x \in N(\mA)$, then $\mA x = 0 \Rightarrow \mA^T 0 = 0$, which implies that for any $x \in N(\mA)$, the vector $x \in N(\mA^T \mA)$. Therefore, we only need to prove that for any $x \in N(\mA^T \mA)$, the vector $x \in N(\mA)$. We prove this by contradiction.

	Suppose a vector $x \in N(\mA^T \mA^T)$ but $x \notin N(\mA)$. We have
	\[ \mA x = b \neq 0, \quad \mA^T \mA x = 0 \quad \Rightarrow \quad \mA^T b = 0. \]
	By the first equation $b \in C(\mA)$, and by the third equation $ b \in N(\mA^T)$ . This contradicts the theorem~\ref{thm:ortho}, i.e.\ $C(\mA) \perp N(\mA^T)$.

	Next, given $N(\mA^T \mA) = N(\mA)$, the rank of matrix $rank(\mA^T \mA) = n - dim(N(\mA^T \mA)) = n - dim(N(\mA)) = rank(\mA)$.
\end{proof}

\begin{cor}[Invertibility of $\mA^T \mA$]\label{cor:invert}
	If $\mA$ has independent columns, then $\mA^T \mA$ is invertible.
\end{cor}
\begin{proof}
	If $\mA \in \bbR^{m \times n}$ has independent columns, then $rank(\mA) = n$. By the theorem~\ref{thm:aa}, $rank(\mA^T \mA) = rank(\mA) = n$. Since $\mA^T \mA \in \bbR^{n \times n}$ is a square matrix, $\mA^T \mA$ is invertible.
\end{proof}

\subsection{Projections onto subspaces}
\begin{defn}[\textit{Projection and projection matrix}]\label{def:project}
	Consider a vector $x \in \bbR^{m}$ and a matrix $\mA^{m \times n}$ that has independent columns. Suppose a vector $p \in C(\mA)$, such that

	\begin{equation}\label{eq:project}
		(x-p) \perp C(\mA). 
	\end{equation}

	The vector $p$ is the projection of vector $x$ onto the space $C(\mA)$. Since $p \in C(\mA)$, there exists a vector $\hat x$ such that $p = \mA \hat x$. By equation~\eqref{eq:project}, we have
	\[ \mA^T (x - p) = \mA^T(x - \mA \hat x) \quad  \Rightarrow\quad  \hat x = (\mA^T \mA)^{-1} \mA^T x,\quad p = \mA(\mA^T \mA)^{-1} \mA^T x.  \] 
	The matrix $\mP =  \mA(\mA^T \mA)^{-1} \mA^T$ is called projection matrix.
\end{defn}

@@

\begin{cor}[Projection onto $N(\mA^T)$]\label{cor:iprojectm}
	Suppose $\mP$ is a projection matrix in theorem~\ref{thm:projectm} , then $I-\mP$ is also a projection matrix to the left nullspace $N(\mA^T)$. 
\end{cor}

\begin{proof}
	For any $x \in \bbR^m$, we have $x - \mP x \perp C(\mA)$
	\[ x - \mP x \perp C(\mA) \quad \Rightarrow \quad (I-\mP)x \perp C(\mA) \quad \Rightarrow \quad  (I-\mP)x \in N(\mA^T) \text{ and } (x - (I-\mP)x) \perp N(\mA^T). \]
	Therefore, $I-\mP$ is a projection matrix to the left nullspace $N(\mA^T)$.
\end{proof}

\subsection{Projection matrices and least squares}
Given observation vector $y \in \bbR^n$ and the design matrix $\mX \in \bbR^{n \times (k+1)}$, we propose the linear regression model
\[ y = \mX \beta +\epsilon,\]
where $\beta = (\beta_0,\beta_1,...,\beta_k)$ are coefficients of our interests and $\epsilon$ is the noise. The least square estimate of $\beta$ minimizes the loss
\[  \hat \beta_{LS} = \argmin_{\beta \in \bbR^{k+1}} \onorm{y - \mX \beta}^2, \]
where $\onorm{\cdot}$ is the euclidean norm. The vector $\mX 
\hat \beta_{LS}$ can be considered as a the projection of $y$ onto the column space of $\mX$. Therefore, we may use projection tools to solve the least square estimate.

@@

The estimate $\hat \beta_{LS}$ is identical to the estimates solved by other methods  using the derivative.

@@

\begin{defn}[\textit{Orthonormal matrix and orthogonal matrix}]\label{def:orthonom}
	Consider a matrix $\mQ \in \bbR^{m \times n}$. If the columns of $\mQ$ are orthonormal, the matrix $\mQ$ is an orthonormal matrix. If $ m= n$, the square matrix $\mQ$ is a orthogonal matrix.
\end{defn}

@@

\begin{defn}[\textit{Gram-Schimidt Process and QR decomposition}]\label{gram-schimidt}
Consider a matrix $\mA \in \bbR^{m \times n}$ with $rank(\mA) = n$. Gram-Schimidt process finds the orthonormal basis for $C(\mA)$. Let $a_i \in \bbR^{m}, i \in [m]$ be the columns of matrix $\mA$. 
\begin{align}
	&u_1 = a_1, &&\quad e_1 = \frac{u_1}{\onorm{u_1}}\\
	&u_2 = a_2 - \frac{u_1^T a_2}{u_1^T u_1} u_1 ,&&\quad  e_2 = \frac{u_2}{\onorm{u_2}}\\
	&u_3 = a_3 - \frac{u_1^T a_3}{u_1^T u_1} u_1 - \frac{u_2^T a_3}{u_2^T u_2} u_2,&&\quad  e_2 = \frac{u_3}{\onorm{u_3}}\\
	\vdots
\end{align}
 The vectors $e_1,...,e_n$ are orthonormal basis of the $C(\mA)$. By matrix operations, we obtain a decomposition of matrix $\mA$
 \begin{equation}\label{eq:qr}
 	\mA = [a_1,...,a_n] = [e_1,....,e_2] \begin{bmatrix}
 		e_1^T a_1 & e_1^T a_2 & \cdots & e_1^T a_n\\
 		0&e_2^T a_2 & \cdots & e_2^T a_n\\
 		\vdots &\vdots & & \vdots \\
 		0 &0&\cdots & e_n^T a_n
 	\end{bmatrix} = \mQ \mR
 \end{equation}
	where $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix and $\mR \in \bbR^{n \times n}$ is a upper triangular matrix. We call the matrix decomposition as equation~\eqref{eq:qr} QR decomposition.
\end{defn}

@@

\subsection{Properties of determinants}
We give ten properties of determinants. The last seven properties are deduced by first three basic properties
\begin{enumerate}
	\item\label{prop1} The determinant of identity matrix is equal to 1, i.e.\ $\det(\mI) = 1$.
	\item\label{prop2} Exchanging two rows of a matrix reverses the sign of the matrix's determinant. Hence, an odd number of row exchanges reverse the sign of the determinant while an even number of row exchanges maintain the sign of the determinant.
	\item \begin{enumerate} 
	\item[(a)]\label{prop3a} If one row of the matrix is multiplied by a constant $t$, the determinant of the matrix is multiplied by $t$.
	\begin{align}
		\begin{vmatrix}
			ta&tb\\c &d 
		\end{vmatrix} = t \begin{vmatrix}
			a&b\\c &d
			\end{vmatrix}
	\end{align} 
	\item[(b)]\label{prop3b} The determinant is linearly additive on the rows of the matrix.
	\begin{align}
		\begin{vmatrix}
			a+a'&b+b'\\c &d 
		\end{vmatrix} =  \begin{vmatrix}
			a&b\\c &d
			\end{vmatrix} +  \begin{vmatrix}
			a'&b'\\c &d
			\end{vmatrix} 
	\end{align}
	\end{enumerate}

	\item \label{propidenticalrow} If two rows of a matrix are equal, the determinant of the matrix is 0.  
	\begin{proof}
		Suppose $\mA$ has two identical rows $a_i, a_j$. Exchanging $a_i, a_j$, the exchanged matrix $\mA' = \mA$. By property~\ref{prop2}, $\det(\mA) = -\det(\mA') = -\det(\mA)$. Therefore, $\det(\mA) = 0$.
	\end{proof}

	\item \label{propelim} For a square matrix $\mA \in \bbR^{n \times n}$, subtracting $t$ times of the $i$-th row from the $j$-th row does not change the determinant of $\mA$, where $i \neq j$.
	\begin{proof}
		By property~\ref{prop3a}, property~\ref{prop3b}, and property~\ref{propidenticalrow}, 
	\begin{align}
		 \begin{vmatrix}
			a & b \\ c - ta&d -tb
		\end{vmatrix} = \begin{vmatrix}
			a & b \\ c &d
		\end{vmatrix} -t \begin{vmatrix}
			a & b \\ a& b
		\end{vmatrix}  = \begin{vmatrix}
			a & b \\ c &d.
		\end{vmatrix}
	\end{align}
	The proof for higher-dimension matrix is similar.
	\end{proof}

	\item\label{prop0row} If the matrix $\mA$ has a row that is all zeros, the determinant of $\mA$ is 0.
	\begin{proof}
		By property~\ref{prop3a}, letting $t = 0$ leads to property~\ref{prop0row}.
	\end{proof}

	\item\label{propdiag} Suppose $\mA \in \bbR^{n \times n}$ is a triangular matrix with diagonal elements $d_1,...,d_n$. The determinant of $\mA$ is equal to the product $d_1 ...d_n$.

	\begin{proof}
		By property~\ref{propelim}, eliminating the triangular matrix $\mA$ to a diagonal matrix does not change the determinant of $\mA$. By property~\ref{prop3a} and \ref{prop1}, $\det(\mA) = d_1 ...d_n \det(\mI_n) = d_1 ...d_n $. 
	\end{proof}

@@

	\begin{proof}
		If $\mA$ is a singular matrix, the reduced row echelon form of $\mA$, RREF($\mA$), has a row with all 0 elements. By property~\ref{propelim} and \ref{prop0row}, the determinant of $\mA$ is 0.
	\end{proof}

@@

	\begin{proof}
		If at least one of $\mA, \mB$ is singular, the product $\mA \mB$ is also singular. Then, $\det(\mA \mB) = \det(\mA) \det(\mB) = 0$. If both $\mA, \mB$ are invertible, $\mA, \mB$ can be eliminated to diagonal matrices $\mQ_{A}, \mQ_{B}$ by QR decomposition. Suppose $a_1,...,a_n$ and $b_1,...,b_n$ are diagonal elements for $\mQ_A$ and $\mQ_B$ respectively. By property~\ref{propdiag}, $\det(AB) = \prod_{i=1}^n a_i b_i = \det(\mA) \det(\mB)$. 
	\end{proof}

@@

\subsection{Determinant computation}
\begin{prop}[Big formula for computing determinant]
	Consider a square matrix $\mA = \entry{a_{ij}} \in \bbR^{n \times n}$. The big formula for computing the determinant of $\mA$ states below.
	\[ \det(\mA) =  \sum_{(\alpha_1,...,\alpha_n) \in \tP} (-1)^{N(\alpha_1,...,\alpha_n)} a_{1,\alpha_1} ... a_{n \alpha_n},  \]
	where $(\alpha_1, \alpha_2, ...,\alpha_n)$ is the permutation of $(1,2,...,n)$ and $\tP$ is the collection of all possible  $(\alpha_1,...,\alpha_n)$, and $N(\alpha_1,...,\alpha_n)$ is the number of element exchanges from $(\alpha_1,\alpha_2,...,\alpha_n)$  to $(1,2,...,n)$. 
\end{prop}

\begin{proof}
	For a square matrix $\mA = \entry{a_ij} \in \bbR^{n \times n}$, we decompose the  matrix $\mA$ by the summation of $n^n$ matrices, in which there is only one entry comes from $\mA$ in each row and other entries are 0. By property~\ref{prop3b}, the determinant of $\mA$ is equal to the summation of the determinants of these $n^n$ matrices. Take a 3-by-3 square matrix as an example.
	\begin{align}
		\begin{vmatrix}
			a_{11} & a_{12} & a_{13}\\
			a_{21} & a_{22} & a_{23}\\
			a_{31} & a_{32} & a_{33}
		\end{vmatrix} = \begin{vmatrix}
			a_{11} & 0& 0\\
			a_{21} & 0 & 0\\
			a_{31} &0 & 0
		\end{vmatrix} + \begin{vmatrix}
			a_{11} & 0& 0\\
			0 & a_{22} & 0\\
			a_{31} &0 & 0
		\end{vmatrix} + \cdots +\begin{vmatrix}
			0 & 0& a_{13}\\
			0& 0 & a_{23}\\
			 0&0 & a_{33}
		\end{vmatrix} \overset{\Delta}{=} \mA_1 + \mA_2 +\cdots + \mA_{27}.
	\end{align}
	For the square matrix $\mA$,  there are $n!$ non-singular matrices among the $n^n$ matrices. The number of non-singular matrices coincides with the size of $\tP$. Intuitively, there are $n$ ways to choose an element from the first row, after which there are only $n-1$ ways to choose an element from the second row to avoid the zero determinant. Therefore, we have $n \times (n-1) \times (n-2) \times \cdots \times 2 = n!$ non-singular matrices.

	By property~\ref{propdiag} and \ref{prop2}, the determinant for the non-singular matrix follows the formula \[(-1)^{N(\alpha_1,...,\alpha_n)}  a_{1,\alpha_1} ... a_{n \alpha_n},\]
	where $(\alpha_1, \alpha_2, ...,\alpha_n)$ is the permutation of $(1,2,...,n)$ and $\tP$ is the collection of all possible  $(\alpha_1,...,\alpha_n)$, and $N(\alpha_1,...,\alpha_n)$ is the number of element exchanges from $(\alpha_1,\alpha_2,...,\alpha_n)$  to $(1,2,...,n)$. 
\end{proof}

@@

\begin{example}[Tridiagonal matrix]
	One example usage of cofactor formula is computing the determinant of \textit{tridiagonal matrix}. A tridiagonal matrix is a matrix in which only non-zero elements lie on or adjacent to the diagonal. Let $\mT_n \in \bbR^{n \times n}$ denote the tridiagonal matrix of 1's, i.e.\
\begin{align}
	\mT_n = \begin{bmatrix}
		1 &1&0 &\cdots & 0 &0\\
		1&1&1&\cdots &0&0\\
		0&1&1&\cdots &0&0\\
		\vdots &\vdots &\vdots &&\vdots &\vdots \\
		0&0&0&\cdots&1&1
	\end{bmatrix}_{n \times n}
\end{align}

@@

\subsection{Inverse matrices}
Previously, we use Gauss-Jordan elimination to obtain the inverse matrix of the invertible matrix. Here, we apply cofactor formula for computing the inverse matrix.

@@

\begin{prop}[Determinants and volumes]
The absolute determinant of a square matrix $\mA$ is the volume of parallelepiped determined by the rows of $\mA$, i.e.\
	\[ |\det(\mA)| = vol(P), \]
	where $P$ is the parallelepiped determined by the rows of $\mA$.
\end{prop}

>>>>

\subsection{Orthogonal vectors and subspaces}
\begin{defn}[\textit{Orthogonal vectors}]\label{def:vortho}
	Consider two vectors $x,y \in \bbR^{n}$. The vectors $x$ and $y$ are orthogonal,  denoted $x \perp y$, \textbf{iff} $x^T y = y^T x = 0$.
\end{defn}

\begin{defn}[\textit{Orthogonal subspaces}]\label{def:sortho}
	Consider two subspaces $S,T$. The subspaces $S$ and $T$ are orthogonal,  denoted $S \perp T$, \textbf{iff} for any $s \in S$ and $ t \in T$, $s^T t = t^T s = 0$.
\end{defn}

Given a matrix $\mA \in \bbR^{m \times n}$, there are four subspaces related to $\mA$: column space $C(\mA)$, row space $C(\mA^T)$, nullspace $N(\mA)$, and left nullspace $N(\mA^T)$. Suppose the rank of $\mA$ is $rank(\mA) = r$, the dimensions of these subspaces are: 
\[ dim(C(\mA)) = dim(C(\mA^T)) = r,\quad  dim(N(A)) = n-r,\quad dim(N(\mA^T)) = m-r. \]

\begin{thm}[Orthogonality of matrix subspaces]\label{thm:ortho}
	Consider a matrix $\mA \in \bbR^{m \times n}$. The row space $C(\mA^T)$ and the nullspace $N(\mA)$ are orthogonal; the column space $C(\mA)$ and the left nullspace $N(\mA^T)$ are orthogonal, i.e.\
	\[ C(\mA^T) \perp N(\mA); \quad C(\mA) \perp N(\mA^T).  \]
\end{thm}

\begin{proof}
	Consider the matrix $\mA \in \bbR^{m \times n}$. For any vector $x \in N(\mA)$, we have $\mA x = 0$. 
	\begin{align}
		\mA x  = \begin{bmatrix}
			a_1^T x \\ \vdots \\ a_m^T x 
		\end{bmatrix} = \begin{bmatrix}
			0\\ \vdots \\0
		\end{bmatrix},
	\end{align}
	where $a_i \in \bbR^n, i \in [m]$ is the $i$-th  row of $\mA$. By definition~\ref{def:vortho}, $x$ is orthogonal to the rows in matrix $\mA$. For any vector $v \in C(\mA^T)$, $v$ is a linear combination of the rows $v = c_1 a_1 + ...+c_m a_m$, where $c_i, i\in [m]$ are constants. Multiplying vectors $v$ and $x$, we have
	\[  v^T x =  c_1 a_1^T x + \cdots + c_m a_m^T x = 0.  \]
	Therefore, $v \perp x$, and $N(\mA) \perp C(\mA^T)$.

@@

\begin{thm}[Relationship between $\mA^T \mA$ and $\mA$]\label{thm:aa}
	Consider a matrix $\mA \in \bbR^{m \times n}$. We have
	\begin{equation}
		N(\mA^T \mA) = N(\mA) ; \quad rank(\mA^T \mA) = rank(\mA).
	\end{equation}
\end{thm}
\begin{proof}
	Consider a matrix $\mA \in \bbR^{m \times n}$. First, we prove that $N(\mA^T \mA) = N(\mA)$.

	On one hand, if $x \in N(\mA)$, then $\mA x = 0 \Rightarrow \mA^T  \mA x = \mA^T 0 = 0$. Therefore, for any $x \in N(\mA)$, the vector $x \in N(\mA^T \mA)$.

	On the other hand, we  prove that for any $x \in N(\mA^T \mA)$ the vector $x \in N(\mA)$ by contradiction.

	Suppose there is a vector $x \in N(\mA^T \mA^T)$ but $x \notin N(\mA)$. We have
	\begin{equation}\label{eq:aaa}
		\mA x = b \neq 0, \quad \mA^T \mA x = 0 \quad \Rightarrow \quad \mA^T b = 0.
	\end{equation} 
	By the first equality in \eqref{eq:aaa}, $b \in C(\mA)$, and by the third equation in \eqref{eq:aaa}, $ b \in N(\mA^T)$. This contradicts the theorem~\ref{thm:ortho}, $C(\mA) \perp N(\mA^T)$. Therefore, for any $x \in N(\mA^T \mA)$, the vector $x \in N(\mA)$.

	Next, given $N(\mA^T \mA) = N(\mA)$, we have $rank(\mA^T \mA) = n - dim(N(\mA^T \mA)) = n - dim(N(\mA)) = rank(\mA)$.
\end{proof}

\begin{cor}[Invertibility of $\mA^T \mA$]\label{cor:invert}
	If $\mA$ has independent columns, then $\mA^T \mA$ is invertible.
\end{cor}
\begin{proof}
	If $\mA \in \bbR^{m \times n}$ has independent columns, then $rank(\mA) = n$. By theorem~\ref{thm:aa}, $rank(\mA^T \mA) = rank(\mA) = n$. Since $\mA^T \mA \in \bbR^{n \times n}$ is a square matrix, $\mA^T \mA$ is invertible.
\end{proof}

\subsection{Projections onto subspaces}
\begin{defn}[\textit{Projection and projection matrix}]\label{def:project}
	Given a vector $x \in \bbR^{n}$ and a matrix $\mA^{m \times n}$ with independent columns, the vector $p \in C(\mA)$, such that
	\begin{equation}\label{eq:project}
		(x-p) \perp C(\mA),
	\end{equation}

	is the projection of vector $x$ onto the space $C(\mA)$. Since $p \in C(\mA)$, there exists a vector $\hat x$ such that $p = \mA \hat x$. By equation~\eqref{eq:project}, we have
	\[ \mA^T (x - p) = \mA^T(x - \mA \hat x) = 0  \quad  \Rightarrow\quad  \hat x = (\mA^T \mA)^{-1} \mA^T x,\quad p = \mA(\mA^T \mA)^{-1} \mA^T x.  \] 
	The matrix $\mP \overset{\Delta}{=}   \mA(\mA^T \mA)^{-1} \mA^T$ is called a projection matrix.
\end{defn}

@@

\begin{cor}[Projection onto $N(\mA^T)$]\label{cor:iprojectm}
	Suppose $\mP$ is a projection matrix in theorem~\ref{thm:projectm}, then $\mI-\mP$ is also a projection matrix to the left nullspace $N(\mA^T)$. 
\end{cor}

\begin{proof}
	For any $x \in \bbR^n$, we have
	\[ x - \mP x \perp C(\mA) \quad \Rightarrow \quad (I-\mP)x \perp C(\mA) \quad \Rightarrow \quad  (\mI-\mP)x \in N(\mA^T) \text{ and } (x - (\mI-\mP)x) \perp N(\mA^T). \]
	Therefore, $\mI-\mP$ is a projection matrix to the left nullspace $N(\mA^T)$.
\end{proof}

\subsection{Projection matrices and least squares}
Given an observation vector $y \in \bbR^n$ and a design matrix $\mX \in \bbR^{n \times (k+1)}$, we propose the linear regression model
\[ y = \mX \beta +\epsilon,\]
where $\beta = (\beta_0,\beta_1,...,\beta_k)$ are regression coefficients and $\epsilon$ is the noise. The least square estimate of $\beta$ is the minimizer of the loss, i.e.\
\[  \hat \beta_{LS} = \argmin_{\beta \in \bbR^{k+1}} \onorm{y - \mX \beta}^2, \]
where $\onorm{\cdot}$ is the euclidean norm. The vector $\mX 
\hat \beta_{LS}$ can be considered as a the projection of $y$ onto the column space of $\mX$, which minimizes the distance from $y$ to the column space $C(\mX)$. Therefore, we may use projection tools to solve the least square estimate.

@@

The estimate $\hat \beta_{LS}$ consists with the estimates solved by using the derivative.

@@

\begin{defn}[\textit{Orthonormal matrix and orthogonal matrix}]\label{def:orthonom}
	Consider a matrix $\mQ \in \bbR^{m \times n}$. If the columns of $\mQ$ are orthonormal, the matrix $\mQ$ is an orthonormal matrix. If $ m= n$, the square matrix $\mQ$ is an orthogonal matrix.
\end{defn}

@@

\begin{defn}[\textit{Gram-Schimidt Process and QR decomposition}]\label{gram-schimidt}
Consider a matrix $\mA \in \bbR^{m \times n}$ with $rank(\mA) = n$. Gram-Schimidt process finds the orthonormal basis for $C(\mA)$. Let $a_i \in \bbR^{m}, i \in [m]$ be the columns of matrix $\mA$. 
\begin{align}
	&u_1 = a_1, &&\quad e_1 = \frac{u_1}{\onorm{u_1}}\\
	&u_2 = a_2 - \frac{u_1^T a_2}{u_1^T u_1} u_1 ,&&\quad  e_2 = \frac{u_2}{\onorm{u_2}}\\
	&u_3 = a_3 - \frac{u_1^T a_3}{u_1^T u_1} u_1 - \frac{u_2^T a_3}{u_2^T u_2} u_2,&&\quad  e_2 = \frac{u_3}{\onorm{u_3}}\\
	\vdots
\end{align}
 The vectors $e_1,...,e_n$ are orthonormal basis of the $C(\mA)$. By matrix operations, we obtain a decomposition of matrix $\mA$
 \begin{equation}\label{eq:qr}
 	\mA = [a_1,...,a_n] = [e_1,....,e_2] \begin{bmatrix}
 		e_1^T a_1 & e_1^T a_2 & \cdots & e_1^T a_n\\
 		0&e_2^T a_2 & \cdots & e_2^T a_n\\
 		\vdots &\vdots & & \vdots \\
 		0 &0&\cdots & e_n^T a_n
 	\end{bmatrix} = \mQ \mR
 \end{equation}
	where $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix and $\mR \in \bbR^{n \times n}$ is an upper triangular matrix. We call the matrix decomposition in equation~\eqref{eq:qr} QR decomposition.
\end{defn}

@@

\subsection{Properties of determinants}
We give ten properties of determinants. The last seven properties are deduced by first three basic properties.
\begin{enumerate}
	\item\label{prop1} The determinant of identity matrix is equal to 1, i.e.\ $\det(\mI) = 1$.
	\item\label{prop2} Exchanging two rows of a matrix reverses the sign of the matrix's determinant. Hence, an odd number of row exchanges reverse the sign of the determinant while an even number of row exchanges maintain the sign of the determinant.
	\item\label{prop3} \begin{enumerate} 
	\item[(a)] If one row of the matrix is multiplied by a constant $t$, the determinant of the matrix is multiplied by $t$.
	\begin{align}
		\begin{vmatrix}
			ta&tb\\c &d 
		\end{vmatrix} = t \begin{vmatrix}
			a&b\\c &d
			\end{vmatrix}
	\end{align} 
	\item[(b)] The determinant is linearly additive on the rows of the matrix.
	\begin{align}
		\begin{vmatrix}
			a+a'&b+b'\\c &d 
		\end{vmatrix} =  \begin{vmatrix}
			a&b\\c &d
			\end{vmatrix} +  \begin{vmatrix}
			a'&b'\\c &d
			\end{vmatrix} 
	\end{align}
	\end{enumerate}

	\item \label{propidenticalrow} If two rows of a matrix are equal, the determinant of the matrix is 0.  
	\begin{proof}
		Suppose $\mA$ has two identical rows $a_i, a_j$. The matrix after exchanging $a_i, a_j$, $\mA'$, is the same as the original matrix $\mA$. By property~\ref{prop2}, $\det(\mA) = -\det(\mA') = -\det(\mA)$. Therefore, $\det(\mA) = 0$.
	\end{proof}

	\item \label{propelim} For a square matrix $\mA \in \bbR^{n \times n}$, subtracting $t$ times of the $i$-th row from the $j$-th row does not change the determinant of $\mA$, where $i \neq j$.
	\begin{proof}
		By property~\ref{prop3} and property~\ref{propidenticalrow}, 
	\begin{align}
		 \begin{vmatrix}
			a & b \\ c - ta&d -tb
		\end{vmatrix} = \begin{vmatrix}
			a & b \\ c &d
		\end{vmatrix} -t \begin{vmatrix}
			a & b \\ a& b
		\end{vmatrix}  = \begin{vmatrix}
			a & b \\ c &d.
		\end{vmatrix}
	\end{align}
	The proof for higher-dimension matrix is similar.
	\end{proof}

	\item\label{prop0row} If the matrix $\mA$ has a row that is all zeros, the determinant of $\mA$ is 0.
	\begin{proof}
		By property~\ref{prop3}, letting $t = 0$ leads to property~\ref{prop0row}.
	\end{proof}

	\item\label{propdiag} Suppose $\mA \in \bbR^{n \times n}$ is a triangular matrix with diagonal elements $d_1,...,d_n$. Then, the determinant $\det(\mA) = \prod_{i = 1}^n d_i$. 
	\begin{proof}
		By property~\ref{propelim}, eliminating the triangular matrix $\mA$ to a diagonal matrix does not change the determinant of $\mA$. By property~\ref{prop3} and \ref{prop1}, $\det(\mA) = d_1 ...d_n \det(\mI_n) = d_1 ...d_n $. 
	\end{proof}

@@

	\begin{proof}
		If $\mA$ is a singular matrix, the reduced row echelon form of $\mA$, RREF($\mA$), has a row with all 0 entries. By property~\ref{propelim} and \ref{prop0row}, the determinant of $\mA$ is 0.
	\end{proof}

@@

	\begin{proof}
		If at least one of $\mA, \mB$ is singular, the product $\mA \mB$ is also singular. Then, $\det(\mA \mB) = \det(\mA) \det(\mB) = 0$. If both $\mA, \mB$ are invertible, $\mA, \mB$ can be eliminated to diagonal matrices $\mD_{A}, \mD_{B}$ by elimination. Suppose $a_1,...,a_n$ and $b_1,...,b_n$ are diagonal elements for $\mD_A$ and $\mD_B$ respectively. By property~\ref{propdiag}, $\det(AB) = \prod_{i=1}^n a_i b_i = \det(\mA) \det(\mB)$. 
	\end{proof}

@@

\subsection{Determinant computation}
\begin{prop}[Big formula for computing determinant]
	Consider a square matrix $\mA = \entry{a_{ij}} \in \bbR^{n \times n}$. The big formula for computing the determinant of $\mA$ is
	\[ \det(\mA) =  \sum_{(\alpha_1,...,\alpha_n) \in \tP} (-1)^{N(\alpha_1,...,\alpha_n)} a_{1,\alpha_1} ... a_{n \alpha_n},  \]
	where $(\alpha_1, \alpha_2, ...,\alpha_n)$ is the permutation of $(1,2,...,n)$, $\tP$ is the collection of all possible  $(\alpha_1,...,\alpha_n)$, and $N(\alpha_1,...,\alpha_n)$ is the number of necessary exchanges from $(\alpha_1,\alpha_2,...,\alpha_n)$  to $(1,2,...,n)$. 
\end{prop}

\begin{proof}
	For a square matrix $\mA = \entry{a_ij} \in \bbR^{n \times n}$, we decompose the  matrix $\mA$ by the summation of $n^n$ matrices, in which there is only one entry comes from $\mA$ in each row and other entries are 0. By property~\ref{prop3}, the determinant of $\mA$ is equal to the summation of the determinants of these $n^n$ matrices. Take a 3-by-3 square matrix as an example.
	\begin{align}
		\begin{vmatrix}
			a_{11} & a_{12} & a_{13}\\
			a_{21} & a_{22} & a_{23}\\
			a_{31} & a_{32} & a_{33}
		\end{vmatrix} = \begin{vmatrix}
			a_{11} & 0& 0\\
			a_{21} & 0 & 0\\
			a_{31} &0 & 0
		\end{vmatrix} + \begin{vmatrix}
			a_{11} & 0& 0\\
			0 & a_{22} & 0\\
			a_{31} &0 & 0
		\end{vmatrix} + \cdots +\begin{vmatrix}
			0 & 0& a_{13}\\
			0& 0 & a_{23}\\
			 0&0 & a_{33}
		\end{vmatrix} \overset{\Delta}{=} \mA_1 + \mA_2 +\cdots + \mA_{27}.
	\end{align}
	For the square matrix $\mA$,  there are $n!$ non-singular matrices among the $n^n$ matrices. The number of non-singular matrices coincides with the size of $\tP$. Intuitively, there are $n$ ways to choose an element from the first row, after which there are only $n-1$ ways to choose an element from the second row to avoid the zero determinant. Therefore, we have $n \times (n-1) \times (n-2) \times \cdots \times 2 = n!$ non-singular matrices.

	By property~\ref{propdiag} and \ref{prop2}, the determinant for the non-singular matrix follows the formula \[(-1)^{N(\alpha_1,...,\alpha_n)}  a_{1,\alpha_1} ... a_{n \alpha_n},\]
	where $(\alpha_1, \alpha_2, ...,\alpha_n)$ is the permutation of $(1,2,...,n)$, $\tP$ is the collection of all possible  $(\alpha_1,...,\alpha_n)$, and $N(\alpha_1,...,\alpha_n)$ is the number of necessary exchanges from $(\alpha_1,\alpha_2,...,\alpha_n)$  to $(1,2,...,n)$. 
\end{proof}

@@

\begin{example}[Tridiagonal matrix]
	One example of cofactor formula is computing the determinant of \textit{tridiagonal matrix}. A tridiagonal matrix is a matrix in which only non-zero elements lie on or adjacent to the diagonal. Let $\mT_n \in \bbR^{n \times n}$ denote the tridiagonal matrix of 1's, i.e.\
\begin{align}
	\mT_n = \begin{bmatrix}
		1 &1&0 &\cdots & 0 &0\\
		1&1&1&\cdots &0&0\\
		0&1&1&\cdots &0&0\\
		\vdots &\vdots &\vdots &&\vdots &\vdots \\
		0&0&0&\cdots&1&1
	\end{bmatrix}_{n \times n}
\end{align}

@@

\subsection{Inverse matrices}
Previously, we use Gauss-Jordan elimination to obtain the inverse matrix of an invertible matrix. Here, we apply cofactor formula to compute the inverse matrix.

@@

\begin{prop}[Determinants and volumes]
The absolute determinant of a square matrix $\mA$ is the volume of parallelepiped determined by the rows of $\mA$, i.e.\
	\[ |\det(\mA)| = vol(P(\mA)), \]
	where $P(\mA)$ is the parallelepiped determined by the rows of $\mA$.
\end{prop}
